How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-29 docs/PhysicsPaper/StokesErosion.tex
NM

Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows~\cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows~\cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature, for example, typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plan of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

Our method merges highly-efficient and highly-accurate boundary-integral equation (BIE) methods~\cite{baker1986boundary, moore2007evaluation} with stable interface evolution methods~\cite{hou-low-she1994, Moore2013} to simulate the erosion of dense suspensions of solid bodies in the Stokes flow regime relevant for groundwater-flow applications~\cite{quaife2018boundary, chiu2020viscous}. Originally inspired by related work in the high-Reynolds-number regime~\cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in~\cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies~\cite{chiu2020viscous}. Though the governing fluid-flow equations (Stokes) are linear, the nonlinear feedback between evolving microstructure and flow give rise to highly complex and anisotropic configurations.

Figure~\ref{fig1} shows an example simulation with 80 circular bodies of randomized sizes and positions immersed in a Stokes flow that moves from left to right. A plug flow is imposed on artificial walls to the left and right of each frame, and a no-slip boundary condition is imposed along the top and bottom of each frame. The color depicts the local speed of the flow intervening between bodies. Over time, individual bodies erode in response to the shear stresses induced on them, and heterogeneous material removal creates visible features both at the level of individual bodies and the larger-scale configuration. For example, horizontally-oriented channels are clearly visible in the third and fourth frames, and these contribute to the overall anisotropy of the medium.

% preferentially removing solid material, and further opening the channels. The positive feedback creates the channels that are clearly visible by in final two images.

%Maybe cite: \cite{Rycroft2016}

>>>>

Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows~\cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows~\cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plane of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

Our method merges highly-efficient and highly-accurate boundary-integral equation (BIE) methods~\cite{baker1986boundary, moore2007evaluation} with stable interface evolution methods~\cite{hou-low-she1994, Moore2013} to simulate the erosion of dense suspensions of solid bodies in the Stokes flow regime relevant for groundwater-flow applications~\cite{quaife2018boundary, chiu2020viscous}. Originally inspired by related work in the high-Reynolds-number regime~\cite{Ristroph2012, Moore2013, Huang2015, MooreCPAM2017}, our method is documented, validated, benchmarked in~\cite{quaife2018boundary}, and can simulate the erosion of $O(100)$ solid bodies. The more recent Cauchy reformulation of the BIE and the associated quadrature formulas allow us to resolve points of near contact between bodies, thus enabling high-fidelity simulation of dense suspensions of erodable bodies~\cite{chiu2020viscous}. Though the governing fluid-flow equations (Stokes) are linear, the nonlinear feedback between evolving microstructure and flow gives rise to highly complex and anisotropic configurations.

Figure~\ref{fig1} shows an example simulation with 80 circular bodies of randomized sizes and positions immersed in a Stokes flow moving from left to right. 
Color depicts the local speed of the flow intervening between bodies. Over time, individual bodies erode in response to the shear stresses induced on them, and heterogeneous material removal creates visible features both at the level of individual bodies and the larger-scale configuration. For example, horizontally-oriented channels are clearly visible in the third and fourth frames, and these contribute to the overall anisotropy of the medium.

%A plug flow is imposed on artificial walls to the left and right of each frame, and a no-slip boundary condition is imposed along the top and bottom of each frame.
